Notes: New gaugings in 3d from three-form

### 2023-05-30 Draft_3dH3.tex
Pulling Emanuel's updates.

%%%%%%%% generalised stuff
\newcommand{\gL}{\mathcal{L}}
\newcommand{\gM}{\mathcal{M}}
\newcommand{\tw}{{\cal U}}
\newcommand{\kr}[2]{\delta^{#1}_{#2}}
\newcommand{\cU}{{\cal U}}
\newcommand{\UI}{\left(U^{-1}\right)}
\newcommand{\flt}[1]{\underline{#1}}
\newcommand{\fl}[1]{\ov{#1}}
\newcommand{\cY}{{\cal Y}}
\newcommand{\vg}{\mathring{g}}
\newcommand{\cP}{{\cal P}}
\newcommand{\cV}{{\cal V}}
\newcommand{\cVI}{\left(\cV^{-1}\right)}
\newcommand{\cA}{\mathcal{A}}
\newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}}

@@

	\begin{abstract}
		We use $\EE$ Exceptional Field Theory to construct the consistent truncation of IIB supergravity on $S^3 \times S^3 \times S^1$ to maximal 3-dimensional ${\cal N}=16$ gauged supergravity containing the ${\cal N}=(4,4)$ AdS$_3$ vacuum. We explain how to achieve this by adding a 7-form flux to the $S^1$ reduction of the dyonic $\En{7}$ truncation on $S^3 \times S^3$ previously constructed in the literature. Our truncation Ansatz includes, in addition to the ${\cal N}=(4,4)$ vacuum, a host of moduli breaking some or all of the supersymmetries. We explicitly construct the uplift of a subset of these to construct new supersymmetric and non-supersymmetric AdS$_3$ vacua of IIB string theory, and show how the moduli space of AdS$_3$ vacua of six-dimensional gauged supergravity studied in \cite{Eloy:2021fhc} is compactified upon lifting to 10 dimensions. Along the way, we also derive the form of 3-dimensional ${\cal N}=16$ gauged supergravity in terms of the embedding tensor.
	\end{abstract}
	\vskip 0cm

@@

\EM{To do:
\begin{itemize}
	\item Think about title - mention new AdS$_3$ vacua? Lifting AdS$_3$ vacua?
	\item Move 3-d gauged SUGRA to somewhere else??
	\item Discuss IIA (and IIB -- Jerome's story) T-duals
\end{itemize}
}

\section{Introduction}

% \section{\texorpdfstring{\EE}{E8(8)} ExFT and consistent truncations}

@@

The \EE exceptional field theory, first constructed in ref.~\cite{Hohm:2014fxa}, is an \EE duality-covariant formulation of type II and 11d supergravities. It is defined on a set of $3+248$ coordinates made of three-dimensonal external coordinates $\{x^{\mu}\}$ and internal coordinates $\{Y^{M}\}$ in the 248-dimensional adjoint representation of \EE. The dependance of the fields on these coordinates is constrained by the ``section constraints''\footnote{We use the notation $\otimes$ to indicate that both derivatives may act on different functions.}
\begin{equation} \label{eq:sectionconstraints}
	\begin{cases}
		\eta^{MN}\partial_{M}\otimes\partial_{N} = 0,\\
		f^{MN}{}_{P}\,\partial_{M}\otimes\partial_{N} = 0,\\
		(\mathbb{P}_{3875})_{MN}{}^{KL}\partial_{K}\otimes\partial_{L} = 0,
	\end{cases}
\end{equation}
where $f_{MN}{}^{P}$ are the structure constants of \EE, $\eta_{MN} = (1/60)\,f_{MK}{}^{L}f_{NL}{}^{K}$ its Cartan-Killing metric and $(\mathbb{P}_{3875})_{MN}{}^{KL}$ is the projector on the representation $\mathbf{3875}$:
\begin{equation}
	(\mathbb{P}_{3875})_{MN}{}^{KL} = \frac{1}{7}\, \delta_{(M}{}^{K}\delta_{N)}{}^{L}-\frac{1}{56}\,\eta_{MN}\eta^{KL}-\frac{1}{14}\,f^{P}{}_{M}{}^{(K}f_{PN}{}^{L)}.
\end{equation}
Here and in the following, \EE indices are raised and lowered by the Cartan-Killing metric $\eta_{MN}$. The section constraints~\eqref{eq:sectionconstraints} ensure that the fields depends only on the 7- or 8-dimensional physical internal coordinates embeded in $Y^{M}$.

The theory describes the dynamics of the following bosonic fields:
\begin{equation}
	\left\{g_{\mu\nu}, \M_{MN}, A_{\mu}{}^{M}, B_{\mu\,M}\right\},
\end{equation}
with $g_{\mu\nu}$ the 3-dimensinal external metric, $\M_{MN}$ the generalized metric parametrizing the coset space $\EE/\SO{16}$ and the gauge fields $A_{\mu}{}^{M}$ and $B_{\mu\,M}$. It is a gauge theory, invariant under the generalized Lie derivative of parameters $\Upsilon=(\Lambda^{M},\Sigma_{M})$, whose action on a vector $V^{M}$ of weight $\lambda$ is given by
\begin{equation} \label{eq:genlie}
	{\cal L}_{\Upsilon} V^{M} = \Lambda^{N}\partial_{N}V^{M}-60\,(\mathbb{P}_{248})^{M}{}_{N}{}^{K}{}_{L}\,V^{N}\partial_{K}\Lambda^{L} + \lambda\, V^{M}\partial_{N}\Lambda^{N}+f^{MN}{}_{K}\Sigma_{N}V^{K},
\end{equation}
with $(\mathbb{P}_{248})^{M}{}_{N}{}^{K}{}_{L}=(1/60)\,f^{M}{}_{NP}f^{PK}{}_{L}$ the projector on the adjoint representation. These transformations are well-defined (in particular, they close into an algebra) only if the parameters $\Sigma_{M}$ and the fields $B_{\mu\, M}$ are covariantly constrained: they have to satisfy algebraic constraints similar to eq.~\eqref{eq:sectionconstraints} and be compatibile with the partial derivatives. We require that
\begin{equation} \label{eq:sectionconstraint2}
	\begin{cases}
		\eta^{MN}C_{M}\otimes C'_{N} = 0,\\
		f^{MN}{}_{P}\,C_{M}\otimes C'_{N} = 0,\\
		(\mathbb{P}_{3875})_{MN}{}^{KL}C_{K}\otimes C'_{L} = 0,
	\end{cases}
	\forall\  C_{M}, C'_{M} \in \{\partial_{M}, \Sigma_{M}, B_{\mu\,M}\}.
\end{equation}

\paragraph{}
The bosonic action of \EE exceptional field theory is invariant under the transformations~\eqref{eq:genlie} and has the expression
\begin{equation} \label{eq:ExFTaction}
	S_{\rm ExFT} = \int d^{3}x\,d^{248}Y\,\sqrt{\vert g\vert} \bigg(\widehat{R}+\frac{1}{240}\,D_{\mu}\M_{MN}D^{\mu}\M^{MN}+\mathscr{L}_{\rm int}+\frac{1}{\sqrt{\vert g\vert}}\,\mathscr{L}_{\rm CS}\bigg).
\end{equation}
$\widehat{R}$ is the \EE-covariantised Ricci scalar and the covariant derivative is defined as
\begin{equation}
	D_{\mu} = \partial_{\mu} - {\cal L}_{(A_{\mu}, B_{\mu})}.
\end{equation}
$\mathscr{L}_{\rm int}$ is a potential term depending only on internal derivatives:
\begin{equation} \label{eq:ExFTLint}
	\begin{aligned}
		\mathscr{L}_{\rm int} &= \frac{1}{240}\,\M^{MN}\partial_{M}\M^{KL}\partial_{N}\M_{KL} - \frac{1}{2}\,\M^{MN}\partial_{M}\M^{KL}\partial_{L}\M_{NK} \\
		&\quad -\frac{1}{7200}\,f^{NQ}{}_{P}f^{MS}{}_{R}\,\M^{PK}\partial_{M}\M_{QK}\,\M^{RL}\partial_{N}\M_{SL}\\
		&\quad +\frac{1}{2}\,g^{-1}\partial_{M}g\,\partial_{N}\M^{MN} + \frac{1}{4}\,\M^{MN}\,g^{-2}\partial_{M}g\,\partial_{N}g + \frac{1}{4}\,\M^{MN}\,\partial_{M}g_{\mu\nu}\,\partial_{N}g^{\mu\nu}.
	\end{aligned}
\end{equation}
Finally, $\mathscr{L}_{\rm CS}$ is a Chern-Simons term required to impose the on-shell duality between scalar and vector fields:
\begin{equation}
	\begin{aligned}
		\mathscr{L}_{\rm CS} &= \frac{1}{2}\,\varepsilon^{\mu\nu\rho}\,\bigg(F_{\mu\nu}{}^{M}B_{\rho\,M}-f_{KL}{}^{N}\,\partial_{\mu}A_{\nu}{}^{K}\partial_{N}A_{\rho}{}^{L}-\frac{2}{3}\,f^{N}{}_{KL}\,\partial_{M}\partial_{N}A_{\mu}{}^{K}A_{\nu}{}^{M}A_{\rho}{}^{L}\\
		&\quad -\frac{1}{3}\,f_{MKL}\,f^{KP}{}_{Q}\,f^{LR}{}_{S}\,A_{\mu}{}^{M}\partial_{P}A_{\nu}{}^{Q}\partial_{R}A_{\rho}{}^{S}\bigg),
	\end{aligned}
\end{equation}
where $F_{\mu\nu}{}^{M}$ is the covariant field strength of $A_{\mu}{}^{M}$ (see eq. (2.26) of ref.~\cite{Hohm:2014fxa}).

\section{Consistent truncations to 3-dimensional \texorpdfstring{${\cal N}=16$}{N=16} gauged supergravity}
The \EE exceptional field theory is well suited to the construction of consistent truncation of type~II and 11d supergravities to three-dimensional maximal supergravity. These truncations arise as generalised Scherk-Schwarz reductions, described by an \EE-valued twist matrix $U_{\fl{M}}{}^{M}$ and a scale factor $\rho$. The compactification Ansätze are as follows~\cite{Galli:2022idq}:
\begin{equation} \label{eq:GSSansatz}
	\begin{aligned}
		g_{\mu\nu}(x,Y)&=\rho(Y)^{-2}\,\mathring{g}_{\mu\nu}(x),\\
		\M_{MN}(x,Y)&=U^{\fl{M}}{}_{M}(Y)\,U^{\fl{N}}{}_{N}(Y)\,\M_{\fl{MN}}(x),\\
		A_{\mu}{}^{M}(x,Y)=&\rho(Y)^{-1}\,U_{\fl{M}}{}^{M}(Y)\,\A_{\mu}{}^{\fl{M}}(x),\\
		B_{\mu\,M}(x,Y)&=\Sigma_{\fl{M}\,M}(Y)\,\A_{\mu}{}^{\fl{M}}(x),
	\end{aligned}
\end{equation}
with
\begin{equation}
	\Sigma_{\fl{M}\,M}=\frac{\rho^{-1}}{60}\,f_{\fl{M}}{}^{\fl{PQ}}\,U_{\fl{P}P}\,\partial_{M}U_{\fl{Q}}{}^{P}.
\end{equation}
The fields with flat indices $\fl{M}$ (in the $\mathbf{248}$ representation of \EE) describe the three-dimensional theory; all the dependence on the internal manifold is factored out in $U_{\fl{M}}{}^{M}$ and $\rho$. Note that with the definition of $\Sigma_{\fl{M}\,M}$ the condition~\eqref{eq:sectionconstraint2} is automatically satisfied.

The Ansätze~\eqref{eq:GSSansatz} describe a consistent truncation if the following condition is satisfied:
\begin{equation} \label{eq:GenLeibniz}
	\gL_{\ff{U}_{\fl{M}}} {\cU}_{\fl{N}}{}^M = X_{\fl{M}\fl{N}}{}^{\fl{P}}\, \cU_{\fl{P}}{}^M,
\end{equation}
with $\cU_{\fl{M}}{}^M = \rho^{-1} U_{\fl{M}}{}^M$, $\ff{U}_{\fl{M}} = (\cU_{\fl{M}},\,\Sigma_{\fl{M}})$ and $X_{\fl{M}\fl{N}}{}^{\fl{P}}$ a constant tensor. This condition ensures that the factorised form of the Ansätze~\eqref{eq:GSSansatz} is preserved by the generalized Lie derivative, \textit{e.g.}
\begin{equation} \label{eq:GSSexample}
	\begin{aligned}
		D_{\mu}\M(x,Y) &= U^{\fl{M}}{}_{M}(Y)\,U^{\fl{N}}{}_{N}(Y)\,{\cal D}_{\mu}\M_{\fl{MN}}(x),\\
		{\cal L}_{(A_{\mu},B_{\nu})}A_{\mu}(x,Y) &= \rho(Y)^{-1}\,U_{\fl{M}}{}^{M}(Y)\,\llbracket\A_{\mu},\A_{\nu}\rrbracket^{\fl{M}}(x),
	\end{aligned}
\end{equation}
where
\begin{equation}
	{\cal D}_{\mu}\M_{\fl{MN}} = \partial_{\mu}\M_{\fl{MN}}-2\,\A_{\mu}{}^{\fl{P}}X_{\fl{P}(\fl{M})}{}^{\fl{Q}}\M_{\fl{N})\fl{Q}} \quad {\rm and} \quad \llbracket\A_{\mu},\A_{\nu}\rrbracket^{\fl{M}}=X_{\fl{PQ}}{}^{\fl{M}}\A_{\mu}{}^{\fl{P}}\A_{\nu}^{\fl{Q}}.
\end{equation}

>>>>

%%%%%%%% generalised stuff
\newcommand{\gL}{\mathcal{L}}
\newcommand{\gM}{\mathcal{M}}
\newcommand{\tw}{{\cal U}}
\newcommand{\kr}[2]{\delta^{#1}_{#2}}
\newcommand{\cU}{{\cal U}}
\newcommand{\UI}{\left(U^{-1}\right)}
\newcommand{\flt}[1]{\underline{#1}}
\newcommand{\fl}[1]{\ov{#1}}
\newcommand{\cY}{{\cal Y}}
\newcommand{\vg}{\mathring{g}}
\newcommand{\cP}{{\cal P}}
\newcommand{\cV}{{\cal V}}
\newcommand{\cVI}{\left(\cV^{-1}\right)}
\newcommand{\cA}{\mathcal{A}}
\newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}}
\newcommand{\dg}{\vert g \vert}

@@

	\begin{abstract}
		We use $\EE$ Exceptional Field Theory to construct the consistent truncation of IIB supergravity on $S^3 \times S^3 \times S^1$ to maximal 3-dimensional ${\cal N}=16$ gauged supergravity containing the ${\cal N}=(4,4)$ AdS$_3$ vacuum. We explain how to achieve this by adding a 7-form flux to the $S^1$ reduction of the dyonic $\En{7}$ truncation on $S^3 \times S^3$ previously constructed in the literature. Our truncation Ansatz includes, in addition to the ${\cal N}=(4,4)$ vacuum, a host of moduli breaking some or all of the supersymmetries. We explicitly construct the uplift of a subset of these to construct new supersymmetric and non-supersymmetric AdS$_3$ vacua of IIB string theory, and show how the moduli space of AdS$_3$ vacua of six-dimensional gauged supergravity studied in \cite{Eloy:2021fhc} is compactified upon lifting to 10 dimensions. Along the way, we also derive the form of 3-dimensional ${\cal N}=16$ gauged supergravity in terms of the embedding tensor and rule out a 10-/11-dimensional origin of some 3-dimensional gauged supergravities.
	\end{abstract}
	\vskip 0cm

@@

\EM{To do:
\begin{itemize}
	\item Think about title - mention new AdS$_3$ vacua? Lifting AdS$_3$ vacua?
	\item Check the $\SL{n}$ equations
	\item Move 3-d gauged SUGRA to somewhere else??
	\item Construct 10-parameter uplift
	\item Complex geometry of moduli space?
	\item Compare with classification known in literature, e.g. in IIA (and IIB -- Jerome's story)
\end{itemize}
}

\section{Introduction}
Consistent truncations are a powerful technique that simplifies the dynamics of 10-/11-dimensional supergravity by focusing on a restricted subset of fields. The key principle behind a consistent truncation is to truncate to a subsector of fields, such that all solutions of the truncated theory correspond to solutions of the original 10-/11-dimensional supergravity theory. By considering only a, typically finite, subset of fields, consistent truncations provide a powerful tool to find complicated new 10-/11-dimensional supergravity solutions and to study their deformations. Consistent truncations have proven particularly powerful in the AdS/CFT correspondence, since all well-understood AdS vacua of string theory do not admit scale-separation and, thus, cannot be studied using the usual tools of lower-dimensional effective theories.

Exceptional Field Theory (ExFT) is a reformulation of 10-/11-dimensional supergravity that unifies the metric and flux degrees of freedom and thereby makes manifest an exceptional symmetry group\footnote{For the purposes of this paper, we do not draw a distinction between ExFT and Exceptional Generalised Geometry, since these agree when the ``section condition'' is solved, which we will always assume here.}. Over the last decade, this has proven an extremely useful tool for constructing consistent truncations, leading to a number of new examples preserving various amounts of supersymmetry \cite{Hohm:2014qga,Lee:2014mla,Malek:2015hma,Baguet:2015sma,Baguet:2015iou,Lee:2015xga,Malek:2016bpu,Ciceri:2016dmd,Cassani:2016ncu,Inverso:2016eet,Malek:2017cle,Malek:2017njj,Malek:2018zcz,Malek:2019ucd,Cassani:2019vcl,Malek:2020jsa,Cassani:2020cod}. However, consistent truncations to three dimensions have, until recently \cite{Galli:2022idq}, remained largely unexplored. One reason is that the $\EE$ ExFT requires modifications for its local symmetry structure, i.e. the generalised Lie derivative, to close into an algebra \cite{Hohm:2014fxa,Cederwall:2015ica}. Another is that there are no maximally supersymmetric AdS$_3$ vacua, i.e. preserving 32 supercharges, of string theory, leaving no natural candidate for constructing a consistent truncation with vacua. Indeed, while the consistent truncations of 11-dimensional supergravity on $S^7$ and $S^4$ and of 10-dimensional supergravity on $S^5$ contain maximally supersymmetric AdS vacua and are captured by a universal Ansatz in ExFT \cite{Hohm:2014qga,Lee:2014mla}, the analogous $S^8$ truncation of 11-dimensional supergravity does not exist, while there are two $S^7$ truncations of 10-dimensional supergravity but neither contains a maximally symmetric vacuum\cite{Fischbacher:2003yw,Galli:2022idq}.

Nonetheless, there are half-maximal, i.e. ${\cal N}=(4,4)$, supersymmetric AdS$_3$ vacua of string theory that are extremely intriguing. These are the AdS$_3 \times S^3 \times S^3 \times S^1$ and AdS$_3 \times S^3 \times T^4$ of IIB string theory (realising the ``small'' and ``large'' ${\cal N}=(4,4)$ superconformal symmetries), and, unlike in higher dimensions, can be supported by pure NS-NS flux. Not only does this mean that there are also heterotic versions of these vacua (preserving ${\cal N}=(4,0)$ supersymmetry), but also that they can be readily studied via the string worldsheet CFT \cite{Maldacena:2000hw,Eberhardt:2018ouy,Eberhardt:2019niq,Eberhardt:2019ywk}. Moreover, the string sigma model in these vacua is also integrable \cite{}. Finally, the AdS$_3$/CFT$_2$ correspondence has seen much progress recently, with many new holographic duals being established \cite{}.

Here we will focus on the ${\cal N}=(4,4)$ AdS$_3 \times S^3 \times S^3 \times S^1$ vacuum of IIB string theory and show that it admits a consistent truncation to a maximal gauged supergravity, which was first studied in \cite{Hohm:2005ui}. All vacua of this theory break at least half the supersymmetries, reflecting the absence of a maximally supersymmetric AdS$_3$ vacuum in string theory. Key to constructing this consistent truncation is to add new fluxes to the $S^1$ reduction of the consistent truncation of IIB supergravity on $S^3 \times S^3$ \cite{Inverso:2016eet}.

Using our consistent truncation, we will show that the AdS$_3 \times S^3 \times S^3 \times S^1$ has a rich moduli space of symmetry- and supersymmetry-breaking deformations, including some that are analogous to the ``flat deformations'' studied for AdS$_4 \times S^5 \times S^1$ \cite{Guarino:2020gfe,Guarino:2021hrc,Giambrone:2021zvp,Giambrone:2021wsm}. While some of these deformations preserve some amounts of supersymmetry, others break all supersymmetries, yet, surprisingly, at least a subset of vacua continue to be perturbatively stable within IIB supergravity, similar to \cite{Giambrone:2021wsm}. Our work also provides an uplift of 6-dimensional supergravity on $S^3$ studied in \cite{Eloy:2021fhc} to IIB string theory. As we will show, some deformations that appear non-compact in 6-dimensional supergravity \cite{Eloy:2021fhc} are compactified within the full 10-dimensional supergravity theory.

Another technical result of our work is the derivation of the potential of 3-dimensional gauged supergravity in terms of the embedding tensor, which was previously only expressed in terms of fermion shift matrices. As a by-product, we find that a constraint that must be obeyed by the embedding tensor of any 3-dimensional gauged supergravity that can be uplifted to 10-/11-dimensional supergravity. As we show, this allows us to prove a higher-dimensional origin of some 3-dimensional gaugings.

The outline of our paper is as follows. We begin with a review of $\EE$ ExFT in section \ref{s:Review} and how to derive consistent truncations in this formalism in \ref{s:TruncationReview}. In section \ref{s:Potential} we derive the potential of 3-dimensional gauged supergravity in terms of the embedding tensor and prove the lack of higher-dimensional origin of some 3-dimensional theories. Then, we show how to add fluxes to consistent truncations in \ref{s:AddFlux}, allowing us to construct the consistent truncation of IIB supergravity on $S^3 \times S^3 \times S^1$. Using this truncation, we study the moduli space of the AdS$_3$ vacua in section \ref{s:Moduli}, before concluding in section \ref{s:Conclusions}.

@@

The \EE exceptional field theory, first constructed in ref.~\cite{Hohm:2014fxa}, is an \EE duality-covariant formulation of type II and 11d supergravities. It is defined on a set of $3+248$ coordinates made of three-dimensonal external coordinates $\{x^{\mu}\}$ and internal coordinates $\{Y^{M}\}$ in the 248-dimensional adjoint representation of \EE. The dependance of the fields on these coordinates is constrained by the ``section constraints''\footnote{We use the notation $\otimes$ to indicate that both derivatives may act on different functions.}
\begin{equation} \label{eq:sectionconstraints}
	\begin{cases}
		\eta^{MN}\partial_{M}\otimes\partial_{N} = 0\,,\\
		f^{MN}{}_{P}\,\partial_{M}\otimes\partial_{N} = 0\,,\\
		(\mathbb{P}_{3875})_{MN}{}^{KL}\partial_{K}\otimes\partial_{L} = 0\,,
	\end{cases}
\end{equation}
where $f_{MN}{}^{P}$ are the structure constants of \EE, $\eta_{MN} = \tfrac{1}{60}\,f_{MK}{}^{L}f_{NL}{}^{K}$ its Cartan-Killing metric and $(\mathbb{P}_{3875})_{MN}{}^{KL}$ is the projector on the representation $\mathbf{3875}$:
\begin{equation}
	(\mathbb{P}_{3875})_{MN}{}^{KL} = \frac{1}{7}\, \delta_{(M}{}^{K}\delta_{N)}{}^{L}-\frac{1}{56}\,\eta_{MN}\eta^{KL}-\frac{1}{14}\,f^{P}{}_{M}{}^{(K}f_{PN}{}^{L)}\,.
\end{equation}
Here and in the following, \EE indices are raised and lowered by the Cartan-Killing metric $\eta_{MN}$. The section constraints~\eqref{eq:sectionconstraints} ensure that the fields depends only on the 7- or 8-dimensional physical internal coordinates embeded in $Y^{M}$.

The theory describes the dynamics of the following bosonic fields:
\begin{equation}
	\left\{g_{\mu\nu}, \M_{MN}, A_{\mu}{}^{M}, B_{\mu\,M}\right\}\,,
\end{equation}
with $g_{\mu\nu}$ the 3-dimensinal external metric, $\M_{MN}$ the generalized metric parametrizing the coset space $\EE/\SO{16}$ and the gauge fields $A_{\mu}{}^{M}$ and $B_{\mu\,M}$. It is a gauge theory, invariant under the generalized Lie derivative of parameters $\Upsilon=(\Lambda^{M},\Sigma_{M})$, whose action on a vector $V^{M}$ of weight $\lambda$ is given by
\begin{equation} \label{eq:genlie}
	{\cal L}_{\Upsilon} V^{M} = \Lambda^{N}\partial_{N}V^{M}-60\,(\mathbb{P}_{248})^{M}{}_{N}{}^{K}{}_{L}\,V^{N}\partial_{K}\Lambda^{L} + \lambda\, V^{M}\partial_{N}\Lambda^{N}+f^{MN}{}_{K}\Sigma_{N}V^{K}\,,
\end{equation}
with $(\mathbb{P}_{248})^{M}{}_{N}{}^{K}{}_{L}=(1/60)\,f^{M}{}_{NP}f^{PK}{}_{L}$ the projector on the adjoint representation. These transformations are well-defined (in particular, they close into an algebra) only if the parameters $\Sigma_{M}$ and the fields $B_{\mu\, M}$ are covariantly constrained: they have to satisfy algebraic constraints similar to eq.~\eqref{eq:sectionconstraints} and be compatibile with the partial derivatives. We require that
\begin{equation} \label{eq:sectionconstraint2}
	\begin{cases}
		\eta^{MN}C_{M}\otimes C'_{N} = 0\,,\\
		f^{MN}{}_{P}\,C_{M}\otimes C'_{N} = 0\,,\\
		(\mathbb{P}_{3875})_{MN}{}^{KL}C_{K}\otimes C'_{L} = 0\,,
	\end{cases}
	\forall\  C_{M}, C'_{M} \in \{\partial_{M}, \Sigma_{M}, B_{\mu\,M}\}\,.
\end{equation}

\paragraph{}
The bosonic action of \EE exceptional field theory is invariant under the transformations~\eqref{eq:genlie} and has the expression
\begin{equation} \label{eq:ExFTaction}
	S_{\rm ExFT} = \int d^{3}x\,d^{248}Y\,\sqrt{\vert g\vert} \bigg(\widehat{R}+\frac{1}{240}\,D_{\mu}\M_{MN}D^{\mu}\M^{MN}+\mathscr{L}_{\rm int}+\frac{1}{\sqrt{\vert g\vert}}\,\mathscr{L}_{\rm CS}\bigg)\,.
\end{equation}
$\widehat{R}$ is the \EE-covariantised Ricci scalar and the covariant derivative is defined as
\begin{equation}
	D_{\mu} = \partial_{\mu} - {\cal L}_{(A_{\mu}, B_{\mu})}\,.
\end{equation}
$\mathscr{L}_{\rm int}$ is a potential term depending only on internal derivatives, explicitly
\begin{equation} \label{eq:ExFTLint}
	\begin{aligned}
		\mathscr{L}_{\rm int} &= \frac{1}{240}\,\M^{MN}\partial_{M}\M^{KL}\partial_{N}\M_{KL} - \frac{1}{2}\,\M^{MN}\partial_{M}\M^{KL}\partial_{L}\M_{NK} \\
		&\quad -\frac{1}{7200}\,f^{NQ}{}_{P}f^{MS}{}_{R}\,\M^{PK}\partial_{M}\M_{QK}\,\M^{RL}\partial_{N}\M_{SL}\\
		&\quad +\frac{1}{2}\,g^{-1}\partial_{M}g\,\partial_{N}\M^{MN} + \frac{1}{4}\,\M^{MN}\,g^{-2}\partial_{M}g\,\partial_{N}g + \frac{1}{4}\,\M^{MN}\,\partial_{M}g_{\mu\nu}\,\partial_{N}g^{\mu\nu}\,.
	\end{aligned}
\end{equation}
Finally, $\mathscr{L}_{\rm CS}$ is a Chern-Simons term required to impose the on-shell duality between scalar and vector fields, given by
\begin{equation}
	\begin{aligned}
		\mathscr{L}_{\rm CS} &= \frac{1}{2}\,\varepsilon^{\mu\nu\rho}\,\bigg(F_{\mu\nu}{}^{M}B_{\rho\,M}-f_{KL}{}^{N}\,\partial_{\mu}A_{\nu}{}^{K}\partial_{N}A_{\rho}{}^{L}-\frac{2}{3}\,f^{N}{}_{KL}\,\partial_{M}\partial_{N}A_{\mu}{}^{K}A_{\nu}{}^{M}A_{\rho}{}^{L}\\
		&\quad -\frac{1}{3}\,f_{MKL}\,f^{KP}{}_{Q}\,f^{LR}{}_{S}\,A_{\mu}{}^{M}\partial_{P}A_{\nu}{}^{Q}\partial_{R}A_{\rho}{}^{S}\bigg),
	\end{aligned}
\end{equation}
where $F_{\mu\nu}{}^{M}$ is the covariant field strength of $A_{\mu}{}^{M}$ (see eq. (2.26) of ref.~\cite{Hohm:2014fxa}).

\section{Consistent truncations to 3-dimensional \texorpdfstring{${\cal N}=16$}{N=16} gauged supergravity} \label{s:TruncationReview}
The \EE exceptional field theory is well suited to the construction of consistent truncation of type~II and 11d supergravities to three-dimensional maximal supergravity. These truncations arise as generalised Scherk-Schwarz reductions, described by an \EE-valued twist matrix $U_{M}{}^{\fl{M}}$ and a scale factor $\rho$. The truncation Ansätze are as follows~\cite{Galli:2022idq}:
\begin{equation} \label{eq:GSSansatz}
	\begin{aligned}
		g_{\mu\nu}(x,Y)&=\rho(Y)^{-2}\,\mathring{g}_{\mu\nu}(x)\,,\\
		\M_{MN}(x,Y)&=U_{M}{}^{\fl{M}}(Y)\,U_{N}{}^{\fl{N}}(Y)\,\M_{\fl{MN}}(x)\,,\\
		A_{\mu}{}^{M}(x,Y)=&\rho(Y)^{-1}\,\UI_{\fl{M}}{}^{M}(Y)\,\A_{\mu}{}^{\fl{M}}(x)\,,\\
		B_{\mu\,M}(x,Y)&=\Sigma_{\fl{M}\,M}(Y)\,\A_{\mu}{}^{\fl{M}}(x)\,,
	\end{aligned}
\end{equation}
with
\begin{equation}
	\Sigma_{\fl{M}\,M}=\frac{\rho^{-1}}{60}\,f_{\fl{M}}{}^{\fl{PQ}}\,\UI_{\fl{P}P}\,\partial_{M}\UI_{\fl{Q}}{}^{P}\,.
\end{equation}
The fields with flat indices $\fl{M}$ (in the $\mathbf{248}$ representation of \EE) belong to the three-dimensional theory; all the dependence on the internal manifold is factored out in $U_{\fl{M}}{}^{M}$ and $\rho$. Note that with the definition of $\Sigma_{\fl{M}\,M}$ the condition~\eqref{eq:sectionconstraint2} is automatically satisfied.

The Ansätze~\eqref{eq:GSSansatz} describe a \textit{consistent} truncation if the following condition is satisfied:
\begin{equation} \label{eq:GenLeibniz}
	\gL_{\ff{U}_{\fl{M}}} {\cU}_{\fl{N}}{}^M = X_{\fl{M}\fl{N}}{}^{\fl{P}}\, \cU_{\fl{P}}{}^M\,,
\end{equation}
with $\cU_{\fl{M}}{}^M = \rho^{-1} \UI_{\fl{M}}{}^M$, $\ff{U}_{\fl{M}} = (\cU_{\fl{M}},\,\Sigma_{\fl{M}})$ and $X_{\fl{M}\fl{N}}{}^{\fl{P}}$ a constant tensor. This condition ensures that the factorised form of the Ansätze~\eqref{eq:GSSansatz} is preserved by the generalized Lie derivative, \textit{e.g.}
\begin{equation} \label{eq:GSSexample}
	\begin{aligned}
		D_{\mu}\M(x,Y) &= U_{M}{}^{\fl{M}}(Y)\,U_{N}{}^{\fl{N}}(Y)\,{\cal D}_{\mu}\M_{\fl{MN}}(x)\,,\\
		{\cal L}_{(A_{\mu},B_{\nu})}A_{\mu}(x,Y) &= \rho(Y)^{-1}\,\UI_{\fl{M}}{}^{M}(Y)\,\llbracket\A_{\mu},\A_{\nu}\rrbracket^{\fl{M}}(x)\,,
	\end{aligned}
\end{equation}
where
\begin{equation} \label{eq:3dCovDeriv}
	{\cal D}_{\mu}\M_{\fl{MN}} = \partial_{\mu}\M_{\fl{MN}}-2\,\A_{\mu}{}^{\fl{P}}X_{\fl{P}(\fl{M})}{}^{\fl{Q}}\M_{\fl{N})\fl{Q}}\,, \quad {\rm and} \quad \llbracket\A_{\mu},\A_{\nu}\rrbracket^{\fl{M}}=X_{\fl{PQ}}{}^{\fl{M}}\A_{\mu}{}^{\fl{P}}\A_{\nu}^{\fl{Q}} \,.
\end{equation}
